---
title: "Data Acquisition"
author: "Stuart Hertzog"
date: "May 16, 2018"
output: html_document
---

**The means to acquire data are built into `R`**, which makes it extremely easy to access large pools of publicly-available statistical data pools that have been assembled by governments, research centres, corporations, business associations, and non-profit organisations.

Data can be harvested from:

* **Plain text and delimited files** (`.txt`, `.csv`, `.tsv`)
* **Microsoft Word and Excel files** (`.doc`, `.docx`, `.xls`, `.xlsx`)
* **Data contained within databases** (SQLite, MySQL, Microsoft Access)
* **Data in statistical software** (SAS, SPSS, NCSS, Octave)

The data sources can be located:

  * **on your computer storage**
  * **on a local or linked network**, or
  * **anywhere on the global Internet**

## Built-In `R` Datasets

An `R` installation includes many built-in datasets. These can be listed with the command:

```{r, eval = FALSE}
data()
```

Many packages also contain their own datasets. You can see a comprehensive listing of these with the command:

```{r, eval = FALSE}
data(package = .packages(all.available = TRUE)
```

A packge must be installed and loaded before its dataset is available.


