---
title: "Data Acquisition"
author: "Stuart Hertzog"
date: "May 16, 2018"
output: html_document
---

**Data acquisition tools are built into `R`**, which makes it extremely easy to access large pools of publicly-available statistical data pools that have been assembled by governments, research centres, corporations, business associations, non-profit organisations, and individuals.

Data can be harvested from:

* **Plain text and delimited files** (`.txt`, `.csv`, `.tsv`)
* **Microsoft Word and Excel files** (`.doc`, `.docx`, `.xls`, `.xlsx`)
* **Data contained within databases** (SQLite, MySQL, Microsoft Access)
* **Data in statistical software** (SAS, SPSS, NCSS, Octave)

The data sources can be located:

  * **on your computer storage**
  * **on a local or linked network**, or
  * **anywhere on the global Internet**

## Built-In `R` Datasets

**An `R` installation includes many built-in datasets**. These are contained in the package `datasets` and are available at all times. They can be listed with the command:

```{r, eval = FALSE}
data()
```

## Datasets In Packages

**Many packages also contain their own datasets**. A package must be installed and loaded before its dataset is available. You can see a comprehensive listing of the datasets included in all installed packages with the command:

```{r, eval = FALSE}
data(package = .packages(all.available = TRUE))
```

Packages will be listed in alphabetical order followed by the datasets they contain. You will be surprised to see how many datasets are available for your use! Due to changes in dataset organisation in recent `R` versions, you will likely see in Console:

```
Warning messages:
1: In data(package = .packages(all.available = TRUE)) :
  datasets have been moved from package 'base' to package 'datasets'
2: In data(package = .packages(all.available = TRUE)) :
  datasets have been moved from package 'stats' to package 'datasets'
```

## Importing CSV Data

## Importing Excel Data

## Importing JSON Data

## Data In R Binary Files

## Data From Databases

## Data from Web Sites

## From Other Software


<p class="bottom-navbar"><span class="bottom-navbar-contents"><span class="glyphicon glyphicon-arrow-left"> </span> [<span class="bottom-navbar-link">Part 2 - Fundamentals of R</span>](R_Programming.html)&nbsp;&nbsp; | &nbsp;[<span class="bottom-navbar-link">Learning Resources</span>](R_Learning_Resources.html)&nbsp; <span class="glyphicon glyphicon-arrow-right"></span></span></p>
