---
title: "Fundamentals of R"
author: "by Stuart Hertzog"
date: "March&ndash;May 2018"
---

<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

## Four Fundamentals

The essence of R:

```{r}
R <- c(1:4)
R
```

(See [Vectors](#vectors) later).

**[One]** important difference about R:

* **Vector-based**: R is not a procedural language

**[Two]** reasons to use R for Data Science:

* **Designed for data**: R can manipulate big data sets
* **Graphics Are Graspable**: people understand graphical data

**[Three]**  fundamental principles of R per [John Chambers](https://en.wikipedia.org/wiki/John_Chambers_(statistician)):

* **Objects**: Everything that exists in R is an object
* **Functions**: Everything that happens in R is a function call
* **Interfaces**: to other softwares are an integral part of R

**[Four]** ways of programming R:

* **Command line**: entering R commands in a terminal
* **Source file**: running a set of commands from a saved file
* **R GUI interface**: available for Mac, WIndows, and Linux 
* **Code chunks in RStudio**: allows debugging as you write

## Basic Maths

R has all the basic mathematical functions:

```{r}

1 + 1

```

```{r}

1 + 2 + 3

```

```{r}

3 * 7 * 2

```

```{r}

4 / 3

```

R obeys the standard order of mathematical operations (**PEMDAS**):

1. **P**arentheses ( )
2. **E**xponents ^
3. **M**ultiplication x
4. **D**ivision \
5. **A**ddition +
6. **S**ubtraction -

```{r}

(2 ^ 5) + (2 * 5)

```

The use of white space between operators is recommended.

## Variables

Unlike statically-typed languages such a C++, R does not require variable types to be declared. An R variable can represent any data type or R object, such as a function, result, or graphical plot. R variables can be redeclared.

* Variable names can **contain alphanumeric characters** but not **periods `.` or underscores `_`**
* **They cannot _start_ with a number or underscore**
* Variable names are **case sensitive**

### Assigning variables

R variable assignment operators are `<-` (default) and `=` (acceptable).

```{r}
x <- 2
x
```

```{r}
y = 5
y
```

You can also assign left-to-right with `->`, but variables are not often assigned that way.

```{r}
7 -> z
z
```

Assignment operations can be used successively to assign a value to multiple variables

```{r}
a <- b <- 42
a
b
```

You can also use the built-in `assign` function:

```{r}
assign("q", 4)
q
```

### Removing variables

`rm(variablename)` removes a variable.

```{r}
rm(q)
```

## Data Types

R has four main data types:

* Numeric
* Character (a.k.a Nominal)
* Date
* Logical

You can check the type of variable with `class(variablename`)

```{r}
x <- "eh?"
x
class(x)
```

```{r}
y <- 99
y
class(y)
```

### `Numeric` data types

Numeric data includes both integers and decimals ---  positive, negative, and zero --- similar to `float` or `double` in other languages. A numeric value stored in a variable is automatically assumed to be numeric in R.

You can test whether data is numeric with `is.numeric()`:

```{r}
is.numeric(y)
```

And if it's an integer with ``is.integer()`:

```{r}
is.integer(y)
```

The response of `FALSE` is because to set an integer as a variable you must append the value with `L`:

```{r}
y <- 99L
is.integer(y)
```

R promotes `integers` to `numeric` when needed.

### `Character` data types

R handles Character data in two primary ways: as `character` and as `factor`. They are treated differently:

```{r}
x <- "data"
x
class(x)
```

and

```{r}
y <- factor("data")
y
```

The `levels` are attributes of that factor.

To find the length of a `character` (or `numeric`):

```{r}
nchar(x)
```

This does not work for `factor` data.

### `Date` data types

R has numerous types of dates. `Date` and `POSIXct` are the most useful.

```{r}
date1 <- as.Date("2018-03-28")
date1
class(date1)
as.numeric(date1)
```

and

```{r}
date2 <- as.POSIXct("2018-03-28 10:45")
date2
class(date2)
as.numeric(date2)
```

Using `as.numeric` also changes the underlying type:

```{r}
class(date1)
class(as.numeric(date1))
```

### `Logical` data types

`Logical`s can be either `TRUE` (`T` or `1`) or `FALSE` (`F`or 0). `T` and `F` are not recommended as they are simply shortcuts to `TRUE` and `FALSE` and can be overwritten, causing woe, anguish, mayhem, and rioting. (`TRUE` or `F`?)

Logical data types have a similar test function `is.logical()`:

```{r}
k <- TRUE
class(k)
is.logical(k)
```

## Data Structures

R data structures are containers for data elements:

* **Vectors** -- collections of only **_same-type elements_**
* **Matrices** -- rectangular containers of only **_same-type elements_**
* **Data Frames** -- contain **_many types of vectors_** , all of the same length
* **Arrays** -- Vectors with **_dimensions_** for each **_same-type element_**
* **Lists** -- containers for elements of **_multi-type data types_**

### Vectors

Vectors are the heart of R; it is a **vectorised language**. An R `Vector` is:

> **A collection of elements of the same type**.

**Operations are applied to each element of a vector without the need to loop through them**. This separates R from other programming languages and makes it most suited to manipulation and graphical presentation of data.

**Vectors do not have a dimension**: there is no `column` or `row` vector. Unlike `mathematical vectors` there is no difference between column or row orientation.

#### Creating a vector

Vectors are created with `c`, meaning "combine":

```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8)
x
```

Operations are applied to all elements at once:

```{r}
x + 2
x -3
x * 2
x / 4
x^2
sqrt(x)
```

#### Vector creation shortcuts

```{r}
1:8
8:1
-3:4
4:-3
```

#### Accessing vector elements

Any element of a `Vector` can be directly access using [square brackets] to point to it:

```{r}
x
x[4]
x[8]
```

#### Counting within Vectors

You can check the length of a vector:

```{r}
x
length(x)
y
length(y)
length(x + y)
```

and count the number of charactors in a vector:

```{r}
q <- c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight")
q
nchar(q)
```

#### Combining Vectors

Two vectors of the same or different length can be combined:

##### Vectors of the same length

```{r}
x <- 1:8
x
y <- -3:4
y
x + y
x - y
x * y
x / y
x^y
```

##### Vectors of different lengths

**For two `vectors` of different lengths, the shorter vector is recycled**, and R may issue a warning:

```{r}
x + c(1, 2)
x + c(1, 2, 3)

```

#### Comparison of two Vectors

```{r}
x <- c(1:8)
x
x > 5
y <- c(3:10)
y
x > y
```

The `all()` function tests whether all elements are `TRUE`

```{r}
x <-  10:1
y <-  -4:5
x
y
all(x < y)
```
The `any()` function tests is any element is 'TRUE`:
 
```{r}
any(x < y)
```
 
including vectors, matrices, data frames (similar to datasets), and lists (collections of objects).

#### Factor Vectors

`Factors` are an important concept in R. **`Factors` contain `levels`**, which are the unique values of that `factor` variable.

```{r}
q
qFactor <- as.factor(q)
qFactor
```

Note that the order of `levels`does not matter unless the `ordered` argument is set `TRUE`:

```{r}
factor(x=c("High School", "Doctorate", "Masters", "College"),
         levels=c("High School", "College", "Masters", "Doctorate"),
         ordered=TRUE)
```

### Matrices

A familiar mathematical structure, `matrices` are essential to statistics.

> A `Matrix` is a rectangular structure of rows and columns in which every element is of the same type, often all [numerics](#numeric-data-types).

`Matrics` can be acted upon similarly to `Vectors`, with **PEDMAS**-style element-by-element addition, subtraction, division, and equality.

#### Creating a Matrix

```{r}
A <- matrix(1:12, nrow=3)
A
```

Any element of a `matrix`can be directly accessed using [square bracket] co-ordinates:

```{r}
A[2,3]
A[3,4]
```

#### Dimensions of a Matrix

```{r}
nrow(A)
ncol(A)
dim(A)
```

#### Adding Matrices

```{r}
A
```

```{r}
B <-  matrix(13:24, nrow=3)
B
```

```{r}
A + B
```

#### Multiplying Matrices

```{r}
A * B
```

#### Logical querying

```{r}
A == B
```

#### Naming rows and columns

```{r}
colnames(A) <- c("A1", "A2", "A3", "A4")
rownames(A) <- c("First", "Second", "Third")
A
A["First", "A2"]
A[1,2]
```

Two special `vectors` -- `letters` and `LETTERS` -- create lowercase and UPPERCASE letter named matrix columns or rows:

```{r}
C <- matrix(21:40, nrow=2)
colnames(C) <- LETTERS[1:10]
rownames(C) <- c(letters[1:2])
C
```

### Dataframes

The `data.frame` is perhaps the primary reason for R's growing popularity as a powerful, focussed, and flexible language for use in all aspects of Data Science.

> A `data.frame` is a rectangular collection of [vectors](#vectors), all of which are of the same length but [differing data types](#data-types).

A `Data Frame` looks like an **Excel spreadsheet** in that the data is organised into __columns__ and __rows__. In statistical terms, each column is a _variable_ while each row contains specific _observations_. Similar to a [Matrix](#matrices) only in that it is also rectangular, a `data.frame` is a much more flexible and comprehensive data structure.

#### Creating a Dataframe

Using the existing functions:

```{r}
(x <- 8:1)
(y <- -3:4)
(q <- c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight"))
```

The simplest way of creating a `Dataframe` is with the `data.frame()` function:

```{r}
theDF <- data.frame(x, y, q)
theDF
```

This creates an 8x3 `data.frame` consisting of three `vectors`. Notice that the [data types](#data-types) are included below the column headings.

To assign names to the `vectors`:

```{r}
theDF <- data.frame(First=x, Second=y, Third=q)
theDF
```

To assign names to the rows:

```{r}
rownames(theDF) <- c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight")
theDF
```

#### Examining a Dataframe

The `nrow()`, `ncol()`, `dim()`, `rownames()`, and `names()` functions are available to investigate its properties:

```{r}
(nrow(theDF))
(ncol(theDF))
(dim(theDF))
(rownames(theDF))
(names(theDF))
```

Elements of any `vector` of a `data.frame` can be directly accessed using the `$` or `[row, col]` operators:

```{r}
(theDF$Second)
(theDF[7, 3])
```

To specify an entire row, leave out the column specification, _vice versa_ for specifying an entire column:

```{r}
(theDF[2, ])
(theDF[, 2])
```

To specify more than one row or column, use a `vector` of indices:

```{r}
(theDF[3:5, 2:3])
```

To specify multiple columns by name, use a `character vector` of the column names:

```{r}
(theDF[, c("First", "Third")])
```

To find the `class` of the entire `data.frame`:

```{r}
(class(theDF))
```

or the `class` of any `vector`:

```{r}
(class(theDF$Third))
```

#### Displaying a Dataframe

`data.frames` can be small, large, big, huge, or ginormous, depending on their size. The `head()` and `tail()` functions functions print only the first or last few rows, or the number of rows you set:

```{r}
(head(theDF))
(head(theDF, n=5))
(tail(theDF, n=5))
```


### Arrays

> An `Array` is a multidimensional [Vector](#vectors) whose elements are all the same type, but which also have attributes having dimensions (`dim`) that can also be named (`dimnames`).

#### Creating `Arrays`

To create an `Array`, the first element is the row index, the second the column index, and the remaining elements are for the outer dimensions `row`, `column`, `number of arrays`:

```{r}
theArray <- array(1:12, dim = c(2, 3, 2))
theArray
```

#### Accessing Arrays

Individual elements of an `Array` are accesssed using square brackets similar to a `Vector` but in this case by `[row, column, array #]`.

```{r}
theArray[1, , ]
theArray[2, , ]
theArray[1, , 1]
theArray[1, , 2]
theArray[, , 1]
theArray[, , 2]
```

### Lists

> `Lists` are used to store any number of items of any type: all `numeric` or all `character` vectors, or a mix of them; complete `data.frames`; and even other `lists`.

#### Creating Lists

`Lists` are created with the `list()` function. Each argument to the function becomes an element of the list:

```{r}
list(1, 2, 3)
```

Single-element lists can contain multi-element vectors:

```{r}
list(c(1, 2, 3))
```

Here's a two-element list with the second element a five-element `vector`:

```{r}
list1 <- list(c(1, 2, 3), 3:7)
list1
```

A two-element `list` with the first element an `array`, the second element a ten-element `vector`:

```{r}
list2 <- list(theArray, 1:10)
list2
```

#### Creating Empty Lists

Empty `lists` of a determined length are created using a `vector`:

```{r}
(emptyList <- vector(mode = "list", length = 4))
```

**Note**: Enclosing an expression in round brackets displays the results immediately after execution.

#### Naming Lists

`Lists` can have names, and each element of a `list` can have a unique name

```{r}
names(list2)
```

```{r}
(names(list2) <- c("The Array", "The Vector"))
list2

```

#### Naming List Elements

Names can also be assigned to `list` elements during creation using name-value pairs. This can also include naming the `list` itself:

```{r}
(list3 <- list(theARR=theArray, theVECT=1:10, List3=list2))
```

#### Adding To A List

New elements can be added to a `list` by appending a `numeric` or `named` index that does not yet exist:

```{r}
length(list3)
```

Adding a `numeric` index:

```{r}
list3[[4]] <- 11
length(list3)
list3
```

Adding a `named` index:

```{r}
list3[["AddedElement"]] <- 12:16
length(list3)
list3
```

## Introducing Functions

> R is a functional language, so almost every operation in R involves either creating new functions or accessing functions in package libraries.

**An `R` programmer must choose and load a suitable package, select an appropriate function, and supply the arguments needed to make it work**.

This can be as simple as calling a function against an element:

```{r}
mean (x)
```

More complicated functions require supplying their arguments either in the correct order, or specifying their name with an equals sign.

Either way means knowing the capabilities and requirements of any function.

### Function Help

**If you know the name of a function**, entering a question mark followed by the function
name in the _Console_ pane will display its documentation in the _Viewer_ pane:

```{r}
?mean
```

For help on  binary operator (e.g. `+`, `*`, `==`), surround it with backticks:

```{r}
?`==`
```

**If you are not sure which function to use**, you can search using only part of the name
with `apropos()`:

```{r}
apropos("mea")
```

Construction and use of functions will be detailed later in this Tutorial series.

## Nothing There?

It's rare that a data set is complete in every detail. Some observations will be missing, while others may be accurately reported as being not available. Missing values can be represented in many ways: by a dash `-`, a period `.`, or even the number `99`.

**`R` has two types of missing data: `NA` and `NULL`.**

### NA

**`R` recognises NA as an element of a Vector**.

`is.na()` tests each element of a Vector:

```{r}
z <- c(3, 4, NA, 5, 6, NA)
z
is.na(z)

```

This works with any type of Vector.

#### Removing NAs

Some `Base R` functions return NA even if a single element is NA, for example `mean()`:

```{r}
mean(z)
```

`na.rm=TRUE` removes any NAs, allowing these functions to proceed:

```{r}
mean(z, na.rm=TRUE)
```

### NULL

`NULL` is very Zen: it is nothingness, which means that there isn't even anything missing. `NULL` is atomical and can't exist in a Vector. If used in a Vector, it simply disappears:

```{r}
z <- c(3, 4, NULL, 5, 6)
z
```

The test for `NULL` is `is.null():

```{r}
no <- NULL
is.null(no)
is.null(z)
```

`is.null(z)` returns `FALSE` because the `NULL` within it is not recognised.

## Pipes

**Functions can be chained together using the `magrittr` package**, which introduces **the `%>%` operator**, which for simplicity can be read as **then...**

```{r}
library("magrittr")
x <- 1:12
mean(x)
x %>% mean
```

**Using pipes means that code can be read left-to-right**, which is more natural than thetraditional right-to-left R `<-` operator. **Additional arguments** can be named and included **inside parentheses** after the function call

```{r}
z %>% mean(na.rm=TRUE)
```

## Cheat Sheet

A guide to what we have covered so far (and more) can be found in this PDF:
[Data Science Free - basicR](http://datasciencefree.com/basicR.pdf)


<p class="bottom-navbar"><span class="bottom-navbar-contents"><span class="glyphicon glyphicon-arrow-left"> </span> [<span class="bottom-navbar-link">Part 1 - RStudio Ecosystem</span>](RStudio_Ecosystem.html)&nbsp;&nbsp; | &nbsp;[<span class="bottom-navbar-link">Learning Resources</span>](R_Learning_Resources.html)&nbsp; <span class="glyphicon glyphicon-arrow-right"></span></span></p>
